Networks Matter in International Relations:\\A Latent Network Approach to Dyadic Analysis

### 2017-10-25 paper/2_model.tex
typos from MF

\section{\textbf{Dependencies in Dyadic Data}}
In working with relational data, scholars typically begin working with relational data by structuring it as a set of dyadic observations stacked on top of one another. This makes sense if each observation is independent of the others. For example, a conflict sent from the United States to Japan, is assumed to be independent of any action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While the assumption that most begin with is that each dyadic interaction is taking place in isolation of the others, we know this assumption to be false in theory and in practice. Relational data comes with an explicit structure that in general leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least, when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause. To uncover the underlying structure of relational data, it is helpful to shift towards restructuring dyadic data in the form of a matrix---often referred to as an adjacency matrix---as shown in Figure~\ref{fig:adjMatDeps}. Rows designate the senders of an event and columns the receivers. The cross-sections in this matrix represent the actions that were sent by an actor in the row to those designated in the columns. Thus $y_{ij}$ designates an action $y$, such as a conflictual event or trade flow, that is sent from actor $i$ to actor $j$.

Using the structure of an adjacency matrix, we visualize in Figure~\ref{fig:adjMatDeps} the types of first- and second-order dependencies that can complicate the analysis of relational data in traditional GLMs. The adjacency matrix on the top left highlights a particular row of an adjacency matrix, to illustrate that these values may be more similar to each other than other values in the adjacency matrix because each has a common sender. Homogeneity in interactions involving a common sender also manifest heterogeneity in how active actors are across the network when compared to each other. Thus in most relational datasets (e.g., trade flows, conflict, participation in international organizations, even networks derived from Twitter or Facebook) we often find that there are some actors that are much more active than others \citep{barabasi:reka:1999}. Unless one is able to develop a model that can account for the variety of explanations that may play a role in determining why a particular actor is more active than others, parameter estimates from standard statistical models will be biased.

@@

The presence of these dependencies in relational data points to the fact that there is a complex structure underlying the dyadic events that we observe, and that accounting for this structure is necessary if we are to adequately represent the data generating process. Of course, if one is able to specify each of the nodal, dyadic, and triadic set of attributes that influence interactions then one can be assured that the conditional independence assumption underlying standard approaches will be satisfied. However, it is rarely the case that we can do this even for TSCS data, thus we more than often make modeling decisions to account for that structure. Failing to account for the underlying structure in either TSCS or dyadic data leads to a number of well-known challenges: a) biased estimates of the effect of independent variables, b) uncalibrated confidence intervals, and c) poor predictive performance. Further by ignoring these potential interdependencies, we often ignore substantively interesting features of the problem under study. The study of international relations is founded on the relations among actors. Why ignore the interdependencies that led to the study of IR in the first place?

@@

The AME model combines the social relations regression model (SRRM) to account for nodal and dyadic dependencies and the latent factor model (LFM) for third order dependencies.  details on the SRRM see \citet{li:loken:2002,dorff:minhas:2017}. An earlier version of the LFM used in AME is presented as the general bilinear mixed effects (GBME) model in \citet{hoff:ward:2004}. The GBME model is more limited in the types of dependence patterns that it can capture due to the formulation of the matrix decomposition procedure underlying the LFM. The AME model is specified as follows:

>>>>

\section{\textbf{Dependencies in Dyadic Data}}
In working with relational data, scholars typically begin by structuring these data as a set of dyadic observations stacked on top of one another. This makes sense if each observation is independent of the others. For example, a conflict sent from the United States to Japan, is assumed to be independent of any action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While the assumption that most begin with is that each dyadic interaction is taking place in isolation of the others, we know this assumption to be false in theory and in practice. Relational data comes with an explicit structure that in general leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least, when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause. To uncover the underlying structure of relational data, it is helpful to shift towards restructuring dyadic data in the form of a matrix---often referred to as an adjacency matrix---as shown in Figure~\ref{fig:adjMatDeps}. Rows designate the senders of an event and columns the receivers. The cross-sections in this matrix represent the actions that were sent by an actor in the row to those designated in the columns. Thus $y_{ij}$ designates an action $y$, such as a conflictual event or trade flow, that is sent from actor $i$ to actor $j$.

Using the structure of an adjacency matrix, we visualize in Figure~\ref{fig:adjMatDeps} the types of first- and second-order dependencies that can complicate the analysis of relational data in traditional GLMs. The adjacency matrix on the top left highlights a particular row to illustrate that these values may be more similar to each other than other values because each has a common sender. Interactions involving a common sender also manifest heterogeneity in how active actors are across the network when compared to each other. Thus, in most relational datasets (e.g., trade flows, conflict, participation in international organizations, even networks derived from Twitter or Facebook), we often find that there are some actors that are much more active than others \citep{barabasi:reka:1999}. Unless one is able to develop a model that can account for the variety of explanations that may play a role in determining why a particular actor is more active than others, parameter estimates from standard statistical models will be biased.

@@

The presence of these dependencies in relational data points to the fact that there is a complex structure underlying the dyadic events that we observe, and that accounting for this structure is necessary if we are to adequately represent the data generating process. Of course, if one is able to specify each of the nodal, dyadic, and triadic set of attributes that influence interactions then one can be assured that the conditional independence assumption underlying standard approaches will be satisfied. However, it is rarely the case that we can do this even for TSCS data, thus we more than often make modeling decisions to account for that structure. Failing to account for the underlying structure in either TSCS or dyadic data leads to a number of well-known challenges: a) biased estimates of the effect of independent variables, b) uncalibrated confidence intervals, and c) poor predictive performance. Further by ignoring these potential interdependencies, we often ignore substantively interesting features of the problem. The study of international relations is founded on the relations among actors. Why ignore the interdependencies that led to the study of IR in the first place?

@@

The AME model combines the social relations regression model (SRRM) to account for nodal and dyadic dependencies and the latent factor model (LFM) for third-order dependencies.  For details on the SRRM see \citet{li:loken:2002,dorff:minhas:2017}. An earlier version of the LFM  used in AME is presented as the general bilinear mixed effects (GBME) model in \citet{hoff:ward:2004}. The GBME model is more limited in the types of dependence patterns that it can capture due to the formulation of the matrix decomposition procedure underlying the LFM. The AME model is specified as follows:

### 2017-10-25 paper/3_sim.tex
typos from MF

Next, we estimate the 95\% confidence interval for the three models in each of the simulations and estimate the proportion of times that the true value actually fell in those intervals. The results are summarized in Figure~\ref{fig:ameCalib}, and again we see that the AME approach performs as well as the oracle while the naive approach performs poorly.

>>>>

Next, we estimate the 95\% confidence interval for the three models in each of the simulations and estimate the proportion of times that the true value fell in those intervals. The results are summarized in Figure~\ref{fig:ameCalib}, and again we see that the AME approach performs as well as the oracle, while the naive approach performs poorly by comparison.

### 2017-10-25 paper/4_app.tex
typos from MF

The re-estimation of \citet{weeks:2012} likely has the sharpest divergence between the GLM results and those of the AME Model. In Weeks's initial models,  she finds that machines are less prone to initiate conflict than the reference category, whereas Juntas, Bosses and Strong-men are more conflict-prone, in line with her theoretical expectations. When we look at the posterior distribution of these coefficients in the AME results, we find that none of these values are distinguishable from zero. Similarly, we find less pronounced effects for military capabilities. One explanation for this divergence is the AME model's ability to account for third order effects. Inspection of the multiplicative effects in Figure~\ref{fig:weekscirc} reveals a number of clusters of states which exhibit structural equivalence---in the top left corner we see states like the US, the UK, and Israel who often send conflict to similar targets. There is a cluster of ``rogue states,'' who are receivers of conflict, in the top left (Iran, Iraq, Syria, Libya, North Korea). In the bottom corner we see a cluster of authoritarian senders including Iraq, Russia, Syria, North Korea and China. In general, these clusters have similar governmental types (Iraq, Syria, Libya, and North Korea all fell under the ``boss" category), and similar conflict behavior. In the GLM, which ignores these third order dependencies, many of these results might have been attributed to regime type. Structural equivalence is present even when accounting for nodal characteristics like regime type.  The AME model, on the other hand, shows that this can be specified in terms of the interdependencies captured by the multiplicative effects. In terms of out-of-sample performance, shown in Figure~\ref{fig:weeksauc}, the AME model performs markedly better than the GLM out-of-sample, lending credence to the possibility that these third-order dependencies cause spurious effects for regime type.  In terms of performance, the original Weeks' model does slightly better than a coin-toss overall, and fails to predict more than a couple of events (incorrectly) in almost one million observations.

>>>>

The re-estimation of \citet{weeks:2012} likely has the sharpest divergence between the GLM results and those of the AME Model. In Weeks' initial models,  she finds that machines are less prone to initiate conflict than the reference category, whereas Juntas, Bosses and Strong-men are more conflict-prone, in line with her theoretical expectations. When we look at the posterior distribution of these coefficients in the AME results, we find that none of these values are distinguishable from zero. Similarly, we find less pronounced effects for military capabilities. One explanation for this divergence is the AME model's ability to account for third order effects. Inspection of the multiplicative effects in Figure~\ref{fig:weekscirc} reveals a number of clusters of states which exhibit structural equivalence---in the top left corner we see states like the US, the UK, and Israel who often send conflict to similar targets. There is a cluster of ``rogue states,'' who are receivers of conflict, in the top left (Iran, Iraq, Syria, Libya, North Korea). In the bottom corner we see a cluster of authoritarian senders including Iraq, Russia, Syria, North Korea and China. In general, these clusters have similar governmental types (Iraq, Syria, Libya, and North Korea all fell under the ``boss" category), and similar conflict behavior. In the GLM, which ignores these third order dependencies, many of these results might have been attributed to regime type. Structural equivalence is present even when accounting for nodal characteristics like regime type.  The AME model, on the other hand, shows that this can be specified in terms of the interdependencies captured by the multiplicative effects. In terms of out-of-sample performance, shown in Figure~\ref{fig:weeksauc}, the AME model performs markedly better than the GLM out-of-sample, lending credence to the possibility that these third-order dependencies cause spurious effects for regime type.  In terms of performance, the original Weeks' model does slightly better than a coin-toss overall, and fails to predict more than a couple of events (incorrectly) in almost one million observations.
